The \proglang{R} Package \pkg{forestinventory}: Design-Based Global and Small Area Estimations for Multiphase Forest Inventories

### 2017-08-23 JStatSoft_forestinventory_master-concordance.tex
update Andreas 23.08.2017

\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:%
1 123 1 1 4 1 1 1 4 4 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Introduction.Rnw:ofs 131:%
1 95 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 227:%
139 34 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Main structure and modules of the Package.Rnw:ofs 262:%
1 78 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 341:%
175 13 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Onephase_Estimators and their Application.Rnw:ofs 355:%
1 25 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 381:%
190}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Twophase_Estimators and their Application.Rnw:ofs 382:%
1 124 1 1 2 4 0 1 2 14 1 1 5 7 0 1 2 2 1 1 2 24 0 1 2 4 1 1 2 4 0 1 2 2 1 1 2 1 0 1 5 6 0 1 2 3 1 1 2 8 0 1 2 157 1 1 7 6 0 1 1 26 0 1 2 3 1 1 2 13 0 1 2 3 1 1 6 5 0 1 1 8 0 1 2 5 1 1 6 5 0 1 1 8 0 1 2 7 1 1 2 4 0 1 2 1 1 1 6 5 0 1 1 3 0 1 2 1 1 1 2 11 0 1 2 3 1 1 6 5 0 1 1 8 0 1 2 4 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 911:%
192}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Threephase_Estimators and their Application.Rnw:ofs 912:%
1 118 1 1 2 8 0 1 2 4 1 1 2 1 0 1 1 3 0 1 2 4 1 1 4 1 1 1 6 8 0 1 2 2 1 1 2 27 0 1 2 4 1 1 4 6 0 1 2 2 1 1 2 8 0 1 2 179 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1302:%
194 10 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Calculation of Confidence Intervals.Rnw:ofs 1313:%
1 13 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1327:%
206 8 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Special Cases and Scenarios.Rnw:ofs 1336:%
1 21 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1358:%
216 15 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Analysis and Visualization.Rnw:ofs 1374:%
1 20 1 1 4 6 0 1 2 2 1 1 6 8 0 1 2 2 1 1 7 9 0 1 2 2 1 1 6 8 0 1 2 2 1 1 7 9 0 1 2 4 1 1 5 7 0 1 2 6 1 1 2 4 0 1 2 2 1 1 2 26 0 1 2 16 1 1 4 1 1 1 2 11 0 1 2 18 1 1 5 7 0 1 2 4 1 1 3 5 0 1 2 10 1 1 2 4 0 1 2 36 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1629:%
233 3 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Acknowledgements.Rnw:ofs 1633:%
1 4 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1638:%
238 11 1}

>>>>

\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:%
1 123 1 1 4 1 1 1 4 4 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Introduction.Rnw:ofs 131:%
1 95 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 227:%
139 34 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Main structure and modules of the Package.Rnw:ofs 262:%
1 83 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 346:%
175 13 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Onephase_Estimators and their Application.Rnw:ofs 360:%
1 25 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 386:%
190}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Twophase_Estimators and their Application.Rnw:ofs 387:%
1 124 1 1 2 4 0 1 2 14 1 1 5 7 0 1 2 2 1 1 2 24 0 1 2 4 1 1 2 4 0 1 2 2 1 1 2 1 0 1 5 6 0 1 2 3 1 1 2 8 0 1 2 157 1 1 7 6 0 1 1 26 0 1 2 3 1 1 2 13 0 1 2 3 1 1 6 5 0 1 1 8 0 1 2 5 1 1 6 5 0 1 1 8 0 1 2 7 1 1 2 4 0 1 2 1 1 1 6 5 0 1 1 3 0 1 2 1 1 1 2 11 0 1 2 3 1 1 6 5 0 1 1 8 0 1 2 4 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 916:%
192}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Threephase_Estimators and their Application.Rnw:ofs 917:%
1 119 1 1 2 8 0 1 2 4 1 1 2 1 0 1 1 3 0 1 2 4 1 1 4 1 1 1 6 8 0 1 2 2 1 1 2 27 0 1 2 4 1 1 4 6 0 1 2 2 1 1 2 8 0 1 2 193 1 1 3 2 0 1 1 3 0 1 2 4 1 1 8 11 0 1 3 1 1 1 4 1 1 1 2 12 0 1 2 4 1 1 7 9 0 1 2 2 1 1 2 12 0 1 2 7 1 1 7 10 0 1 3 2 1 1 2 12 0 1 2 6 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1436:%
194 10 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Calculation of Confidence Intervals.Rnw:ofs 1447:%
1 13 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1461:%
206 8 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Special Cases and Scenarios.Rnw:ofs 1470:%
1 21 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1492:%
216 15 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Analysis and Visualization.Rnw:ofs 1508:%
1 20 1 1 4 6 0 1 2 2 1 1 6 8 0 1 2 2 1 1 7 9 0 1 2 2 1 1 6 8 0 1 2 2 1 1 7 9 0 1 2 4 1 1 5 7 0 1 2 6 1 1 2 4 0 1 2 2 1 1 2 26 0 1 2 16 1 1 4 1 1 1 2 11 0 1 2 18 1 1 5 7 0 1 2 4 1 1 3 5 0 1 2 10 1 1 2 4 0 1 2 36 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1763:%
233 3 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Acknowledgements.Rnw:ofs 1767:%
1 4 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1772:%
238 11 1}

### 2017-08-23 rnw/Main structure and modules of the Package.Rnw
update Andreas 23.08.2017

The two-phase or double-sampling estimators use inventory information from \textbf{two} nested samples which are commonly reffered to as \textit{phases} (figure \ref{fig:concmphase_and_sae_a}). The first phase $s_1$ comprises $n_1$ sample locations that each provide a set of explanatory variables described by the column vector $\pmb{Z}(x)\in{\Re^{p}}$ at each point $x \in s_1$. These explanatory variables are derived from auxiliary information that is available in high quantity within the forest area $F$. The second phase $s_2$ constitutes the terrestrial inventory conducted at $n_2$ subsamples of the large phase $s_1$ and provides the value of the target variable, i.e. the local density $Y(x)$ (e.g. the timber volume density per hectare). The set of explanatory variables at each sample location $x \in s_1$ is transformed into a prediction $\hat{Y}(x)$ of $Y(x)$ by the application of an oridinary least square (OLS) regression model. The basic idea of this setup is to boost the sample size by providing a large sample of less precise but cheap information about $Y(x)$ in $x \in s_1$, and to account for the lack of precision by collecting precise but costly information of $Y(x)$ at few sample locations $x \in s_2 \subset s_1$. In the \proglang{R} package \pkg{forestinventory}, estimators for two-phase sampling can be applied with the \code{twophase()} function.

@@

The three-phase or triple-sampling estimators extend the principle of two-phase sampling and use inventory information from \textbf{three} nested samples (\textit{phases}) (figure \ref{fig:concmphase_and_sae_a}). The basic assumption is that auxiliary information is available in two different frequencies, i.e. sample sizes. The phase $s_0$ provides a large number $n_0$ of auxiliary information, whereas the phase $s_1$ provides additional auxiliary information that is only available at $n_1$ subsamples of $s_0$. The terrestrial information is then collected at a subsample $s_2$ of $s_1$ and $s_0$. The basic idea of three-phase sampling is that the additional set of auxiliary information $\pmb{Z}^{(1)}(x)\in{\Re^q}$ adds considerable explanatory power to the set of auxiliary information $\pmb{Z}^{(0)}(x)\in{\Re^p}$ available at all sample locations $x \in s_0$. From that it follows that we can define \textit{two nested} regression models. The \textit{full} set of predictor variables $\pmb{Z}^t(x)=(\pmb{Z}^{(0)t}(x),\pmb{Z}^{(1)t}(x))\in{\Re^{p+q}}$ in the sample $s_1$ can be used to calculate predictions $\hat{Y}_{1}(x)$ of $Y(x)$ at all sample locations $x \in s_1$. The regression model applicable to the $s_1$ phase is thus referred to as the \textit{full model}. Whereas we can only use the subset of auxiliary information $\pmb{Z}^{(0)}(x)$ in the so called \textit{reduced model} at the sample locations $x \in s_0$, it allows for creating a much larger sample of less accurate predictions $\hat{Y}_{0}(x)$. If the gain in model precision between the reduced and the full model as well as the sampling fraction between $s_0$ and $s_1$ is sufficiently large, the three-phase estimator normally leads to a further increase in estimation precision compared to the two-phase estimator. In the \proglang{R} package \pkg{forestinventory}, estimators for three-phase sampling can be applied with the \code{threephase()} function.

@@

We enter the topic of small area estimation if we want to make an estimation for a subregion $G$ of the entire inventory area $F$ (figure \ref{fig:concmphase_and_sae_b}). If the spatial extent of the subregion is small with respect to the terrestrial sample density, the number of terrestrial sample locations within the subregion $n_{2,G}$ will be to small to provide sufficient estimation accuracy. In this case, the application of multiphase estimation procedures can be efficient. However, the terrestrial sample size is too small to fit a separate regression model to them. The idea is then to \textit{borrow strength} from the entire terrestrial sample $s_2$ of $F$ and then to apply this  model to the small area. A potential model bias within $G$ is then corrected for by using the model residuals derived in the small area $G$. If no terrestrial samples fall into a small area $G$ (i.e. $n_{2,G}=0$), this bias correction cannot be applied and the resulting \textit{synthetic} estimates are potentially biased. Both two and three-phase small area estimators for synthetic and non-synthetic estimation can be applied by the \code{twophase()} and \code{threephase()} functions in the \proglang{R} package \pkg{forestinventory}.

@@

The formulation of a suitable regression model constitutes an essential part in the application of multiphase inventories. However, the requirements on the regression model properties differ fundamentally between two major mathematical approaches \cite{mandallaz2008}: In the \textit{model-dependent} (also known as \textit{model-based}) framework, the sample locations \textit{x} are fixed and the observation $Y(x)$ taken at location $x$ is assumed to be a random variable as the realization of a stochastic process (i.e the forest). As a consequence, the regression model of a model-dependent regression estimator has to met the model assumptions in the classical statistical sense, i.e. the observation of the target variable \textit{Y}(\textit{x}) is assumed to be an identically distributed, independent random variable. Moreover, the unbiasedness of the model-dependent estimators substantially depends on the correct specification of the regression model, which means that extensive analyses and adjustments must be conducted to ensure that the model adequately describes the underlying stochastic process. In contrast to the model-dependent approach, the \textit{design-based} (also known as \textit{model-assisted}) approach rests upon the randomization of the sample locations $x$ (i.e. the $x$ are independently and uniformly distributed in the forest), while the forest itself is fixed and $not$ the result of a stochastic process. The observation \textit{Y}(\textit{x}) itself still remains a random variable, but solemnly due to the random sample mechanism. A consequence of this approach is that the regression models used in the design-based regression estimators do no longer have to satisfy any model assumptions in the classical statistical sense (e.g. identical distribution and independence of the observation $Y(x)$, or stationarity assumption in the geostatistical frame), as it is the case in the model-based framework \cite{mandallaz2008}.
% NOTES:
% - we can still describe it a bit better, also with some citations of mcroberts and breidenbach (their md-description)

@@

The following sections describe the mathematical details and the application of the multiphase estimators that are implemented in the \proglang{R} package \pkg{forestinventory}. While \cite{mandallaz2008, mandallaz2013c} provides an extensive derivation of all estimators, we will only provide the mathematical formulas that are actually implemented and applied in the functions of the package. For the sake of brevity, we will restrict to simple sampling and provide the equations for cluster sampling in the Appendix. Special cases occuring under cluster sampling are described in section \ref{sec:speccas_and_scen}.

>>>>

The two-phase or double-sampling estimators use inventory information from \textbf{two} nested samples which are commonly reffered to as \textit{phases} (figure \ref{fig:concmphase_and_sae_a}). The first phase $s_1$ comprises $n_1$ sample locations that each provide a set of explanatory variables described by the column vector $\pmb{Z}(x)\in{\Re^{p}}$ at each point $x \in s_1$. These explanatory variables are derived from auxiliary information that is available in high quantity within the forest area $F$. The second phase $s_2$ constitutes the terrestrial inventory conducted at $n_2$ subsamples of the large phase $s_1$ and provides the value of the target variable, i.e. the local density $Y(x)$ (e.g. the timber volume density per hectare). The set of explanatory variables at each sample location $x \in s_1$ is transformed into a prediction $\hat{Y}(x)$ of $Y(x)$ by the application of an oridinary least square (OLS) regression model. The basic idea of this setup is to boost the sample size by providing a large sample of less precise but cheap information about $Y(x)$ in $x \in s_1$, and to account for the lack of precision by collecting precise but costly information of $Y(x)$ at few sample locations $x \in s_2 \subset s_1$. 
%In the \proglang{R} package \pkg{forestinventory}, estimators for two-phase sampling can be applied with the \code{twophase()} function.

@@

The three-phase or triple-sampling estimators extend the principle of two-phase sampling and use inventory information from \textbf{three} nested samples (\textit{phases}) (figure \ref{fig:concmphase_and_sae_a}). The basic assumption is that auxiliary information is available in two different frequencies, i.e. sample sizes. The phase $s_0$ provides a large number $n_0$ of auxiliary information, whereas the phase $s_1$ provides additional auxiliary information that is only available at $n_1$ subsamples of $s_0$. The terrestrial information is then collected at a subsample $s_2$ of $s_1$ and $s_0$. The basic idea of three-phase sampling is that the additional set of auxiliary information $\pmb{Z}^{(1)}(x)\in{\Re^q}$ adds considerable explanatory power to the set of auxiliary information $\pmb{Z}^{(0)}(x)\in{\Re^p}$ available at all sample locations $x \in s_0$. From that it follows that we can define \textit{two nested} regression models. The \textit{full} set of predictor variables $\pmb{Z}^t(x)=(\pmb{Z}^{(0)t}(x),\pmb{Z}^{(1)t}(x))\in{\Re^{p+q}}$ in the sample $s_1$ can be used to calculate predictions $\hat{Y}_{1}(x)$ of $Y(x)$ at all sample locations $x \in s_1$. The regression model applicable to the $s_1$ phase is thus referred to as the \textit{full model}. Whereas we can only use the subset of auxiliary information $\pmb{Z}^{(0)}(x)$ in the so called \textit{reduced model} at the sample locations $x \in s_0$, it allows for creating a much larger sample of less accurate predictions $\hat{Y}_{0}(x)$. If the gain in model precision between the reduced and the full model as well as the sampling fraction between $s_0$ and $s_1$ is sufficiently large, the three-phase estimator normally leads to a further increase in estimation precision compared to the two-phase estimator. 
%In the \proglang{R} package \pkg{forestinventory}, estimators for three-phase sampling can be applied with the \code{threephase()} function.

@@

We enter the topic of small area estimation if we want to make an estimation for a subregion $G$ of the entire inventory area $F$ (figure \ref{fig:concmphase_and_sae_b}). If the spatial extent of the subregion is small with respect to the terrestrial sample density, the number of terrestrial sample locations within the subregion $n_{2,G}$ will be to small to provide sufficient estimation accuracy. In this case, the application of multiphase estimation procedures can be efficient. However, the terrestrial sample size is too small to fit a separate regression model to them. The idea is then to \textit{borrow strength} from the entire terrestrial sample $s_2$ of $F$ and then to apply this  model to the small area. A potential model bias within $G$ is then corrected for by using the model residuals derived in the small area $G$. If no terrestrial samples fall into a small area $G$ (i.e. $n_{2,G}=0$), this bias correction cannot be applied and the resulting \textit{synthetic} estimates are potentially biased. 
% Both two and three-phase small area estimators for synthetic and non-synthetic estimation can be applied by the \code{twophase()} and \code{threephase()} functions in the \proglang{R} package \pkg{forestinventory}.

@@

The formulation of a suitable regression model constitutes an essential part in the application of multiphase inventories. However, the requirements on the regression model properties differ fundamentally between two major mathematical approaches \cite{mandallaz2008}: In the \textit{model-dependent} (also known as \textit{model-based}) framework, the sample locations \textit{x} are fixed and the observation $Y(x)$ taken at location $x$ is assumed to be a random variable as the realization of a stochastic process (i.e the forest). As a consequence, the regression model of a model-dependent regression estimator has to met the model assumptions in the classical statistical sense, i.e. the observation of the target variable \textit{Y}(\textit{x}) is assumed to be an identically distributed, independent random variable. Moreover, the unbiasedness of the model-dependent estimators substantially depends on the correct specification of the regression model, which means that extensive analyses and adjustments must be conducted to ensure that the model adequately describes the underlying stochastic process. In contrast to the model-dependent approach, the \textit{design-based} (also known as \textit{model-assisted}) approach rests upon the randomization of the sample locations $x$ (i.e. the $x$ are independently and uniformly distributed in the forest), while the forest itself is fixed and \textit{not} the result of a stochastic process. The observation \textit{Y}(\textit{x}) itself still remains a random variable, but solemnly due to the random sample mechanism. A consequence of this approach is that the regression models used in the design-based regression estimators do no longer have to satisfy any model assumptions in the classical statistical sense (e.g. identical distribution and independence of the observation $Y(x)$, or stationarity assumption in the geostatistical frame), as it is the case in the model-based framework \cite{mandallaz2008}. The philosophy of the design-based approach is thus to use prediction models to \textit{improve} the efficiency of the estimators without having to \textit{rely} on their correct specification, which makes them very attractive to be used in official statistics. They are therefore also referred to as \textit{model-assisted}. The randomly distribution of the sample locations, which the design-based inference rests upon, is in practice often replaced by the use of systematic grids due to minimization of travelling cost in the terrestrial survey. However, there is reasonable and empirical evidence that softening this assumption is acceptable for point estimation as well as the variance estimation, which will in most cases by slightly overestimated \citep{mandallaz2008, mandallaz2016}.

% NOTES:
% - we can still describe it a bit better, also with some citations of mcroberts and breidenbach (their md-description)

@@

The following sections describe the mathematical details and the application of the multiphase estimators that are implemented in the \proglang{R} package \pkg{forestinventory}. While \cite{mandallaz2008, mandallaz2013techa, mandallaz2013techb,mandallaz2015tech}
provides an extensive derivation of all estimators, we will restrict to the mathematical formulas that are actually implemented in the package. For the sake of brevity, we will restrict to simple sampling and provide the equations for cluster sampling in the Appendix. Special cases occuring under cluster sampling are described in section \ref{sec:speccas_and_scen}.
